Learning action models with minimal observability

### 2018-08-02 doc/aij18/experiments.tex
Minor changes to experiments

\subsection{Setup}
We evaluate \FAMA on fourteen IPC domains that satisfy the \strips\ requirement~\cite{fox2003pddl2}, all taken from the {\sc planning.domains} repository~\cite{muise2016planning}. Table \ref{tab:domain_features} compiles the features of the domains, all these features affect the size of our compilation. From left to right columns report, for each domain, its number of actions, number of predicates, maximum arity of the actions, and maximum arity of the predicates. %and hence, have an impact on the complexity of the learning task. As can be seen in table \ref{tab:domain_features}, the selected domains range from, such as \emph{blocks} and \emph{visitall}, to complex ones, like \emph{floortile} and \emph{satellite}, with higher number of actions and predicates and with higher arity.

@@

\newpage
Figure~\ref{fig:pspace_quality} reports {\em precision} and {\em recall} for the models learned by our approach in the PSPACE-complete scenario (NO action sequence and NO state trajectory). The {\em precision} and {\em recall} values are now lower and they drop as the amount of input knowledge increases. For the PSPACE-complete scenario, {\em precision} and {\em recall} scores have to be taken with a grain of salt because {\bf pure syntax-based metrics are not informative when reformulations appear}. This is the case of such an under-constrained learning task and we provide experimental evidence of this in Section \ref{minimal}. Further, the score drop reflects an increasing number of timeouts meaning that, no solution is found for several domains within the given time-bound (the worst case complexity of this scenario is larger). Computation time (Figure \ref{fig:pspace_time}) is also higher explained again by this larger number of timeouts.

@@

\subsection{Comparison with \ARMS}
Now we compare the performance of \FAMA to that of \ARMS. \ARMS, as any other previous approaches for learning planning models cannot handle the PSPACE-complete scenario. On the other hand, \ARMS assumes input plan traces with \FO action sequences and \NO state trajectories so here, we will also adopt this assumption.

We define a \emph{degree of observability} $\sigma$ for the state trajectory, ranging from 0\% to 100\%, that measures the probability of observing a literal, and evaluate both approaches as $\sigma$ increases using 5 plan traces as input. When $\sigma = 0$ we have a \NO state trajectory, when $\sigma=100$ we have a \FO state trajectory and all cases in-between correspond to the \PO scenario. Figures \ref{fig:comparison_precision} and \ref{fig:comparison_recall} compare \FAMA and \ARMS in terms of {\em precision} and {\em recall}. Horizontal axes represents the degree of observability, while vertical axes show the average {\em precision} (Figure \ref{fig:comparison_precision}) or {\em recall} (Figure \ref{fig:comparison_recall}). \FAMA dominates in terms of {\em precision} in all cases (the models learned by our approach are 15\% to 24\% more precise than those learned by \ARMS) except for \FO state trajectories. A similar trend is drawn for {\em recall} (Figure \ref{fig:comparison_recall}), where the difference is even larger, meaning that our learned models are more complete. These results evidence the \FAMA capacity to learn better models when few plan traces are available but do not imply that our approach is overall better for NP-complete scenarios (Figure~\ref{fig:np_time} showed that \FAMA cannot handle so much input as \ARMS).

@@

In Table \ref{tab:results_minimum_30_30}, which collects the result of the PSPACE case study with 30\% observability, the scores for some domains are missing. This is the case for {\em floor-tile}, and {\em parking} both considered \emph{puzzle-like} domains and that put a strain on the planners. Interestingly enough, {\em hanoi} also qualifies as a \emph{puzzle-like} domain, and we can see how this is reflected on the computation time. With regards to the learned models, we find that they retain similar level of soundness, while their completeness is lower than in the previous case study. This is specially the case for the preconditions, which have gone from 0.94 recall to 0.54. %This results are difficult to explain, given that in this case study the intermediate states contain more information, but
The explanation for these scores is that input actions are strong constraints that play a key role on how close the learned model are with respect to the GTM. The more these actions are missing the more present is reformulation and pure syntax-based metrics become less relevant.

@@

The effect of missing input actions is more evident in our last case study with \NO action sequences and state trajectories (Table \ref{tab:results_minimum_0_0}). In this case we find solutions for all domains. Search is even less constrained so there are far more possible solutions for this learning task. This wider range of solutions is also evidenced by the kind of the learned models, which in spite of being compliant with the input data, are further from the GTM. Both {\em precision} and {\em recall} decreased to 0.6 and 0.5, respectively. Syntax-based metrics are not appropriate for scenarios with minimal observability as they cannot cope with the \emph{reformulations} that frequently occur in these circumstances.

To illustrate this, Figure \ref{fig:macroaction} shows the PDDL encoding of the action model of the {\tt stack} operator learned from plan traces with \NO action sequences and state trajectories. This learned action model removes a block from on top of another block and puts it down on the table. There are two main differences with respect to the GTM: (1), the learned action is unstacking a block instead of stacking it and (2), the block ends on the table after the action is applied, working as an {\tt unstack\&putdown} \emph{macro-action}. We refer to the first difference as \emph{role swapping} and it happens when there are missing actions in the plan trace. If no actions are present in the input traces, the name of actions becomes meanless (actions are effectively anonymous), meaning that they can interchange their behavior with any other comparable action model. The second issue happens when there are missing states in the input traces.

@@

Table \ref{tab:metric_comparison_100_10} shows the results of the first case study. Looking at the high scores of the syntactic metrics, specially {\em recall}, we can conclude that the learned models are, in fact, fairly similar to the GTM. This supports our conclusion that no reformulation occurs in this case study, which also means that the space of possible solutions is restricted to models close to the GTM. The values for {\em sem-Precision} and {\em sem-Recall} are also high across the table, which is exactly the desired behavior for these metrics given that solutions are close to the GTM. In comparison, both versions of {\em recall} show similar scores, while the difference is bigger for the precision metrics. This means that the semantic version is more lenient towards extra preconditions or effects. This matches what we have observed during the experiments for instance, the learned model for the {\tt move} action of the {\em hanoi} domain, specifies that both the origin and destination disks must be bigger than the one moving, but the GTM contains only one of these preconditions. This learned model is semantically correct but sintactically different from the GTM and hence, penalized by the {\em precision} metric.

Finally Table \ref{tab:metric_comparison_0_0} details the results of the \NO/\NO case study. Contrary to the previous case study, the difference between the syntactic and semantic metrics is larger here. Comparing the scores of both versions, we find that learned models that achieved mediocre scores when using the GTM as reference, are in fact reasonable sound and complete, reaching and overall score of 0.91 for both {\em sem-Precision} and {\em sem-Recall}. This means that the learned models by our approach, despite syntactically different from the GTM, require very few editions to explain the testing traces given as input.

@@

Looking at the results of both case studies we can draw two conclusions with regards to the semantic metrics proposed in this paper. The first one is that, when no reformulation occurs, these metrics behave similarly to their syntactic counterparts, which means they are a good substitute when the GTM is not available. The second conclusion is that {\em sem-Precision} and {\em sem-Recall} are better suited to evaluate reformulated models than the original syntactic metrics, since they contemplate valid solutions outside the GTM that successfully explain the given input data.
%Since we are using {\sc Madagascar}, a satisfying planner, the scores for sem-Precision and sem-Recall are approximated.

%Here we evaluate our approach with learning tasks of the kind $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{T}}$, where the action of the executed plans are not available but the initial and goal states are known. When input plans are not available, the planner must not only compute the action models but also the plans that satisfy the input observations. Table~\ref{tab:results_states} and ~\ref{tab:time_states} summarize the results obtained for this using static predicates and partially specified models. Values for the {\em Zenotravel} and {\em Grid} domains are not reported because {\sc Madagascar} was not able to solve the corresponding planning tasks within a 1000 sec. time bound. The values of \emph{precision} and \emph{recall} are significantly lower than in Table ~\ref{tab:results_plans}. Given that the learning hypothesis space is now fairly under-constrained, actions can be reformulated and still be compliant with the inputs (e.g. the {\em blocksworld} operator {\small\tt stack} can be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa). We tried to minimize this effect with the additional input knowledge (static predicates and partially specified action models) and yet the results are below the scores obtained when learning from labeled plans.

%To give an insight of the actual quality of the learned models, we defined a method for computing {\em Precision} and {\em Recall} that is robust to the mentioned model {\em reformulations}. Precision and recall are often combined using the {\em harmonic mean}. This expression, called the {\em F-measure} or the balanced {\em F-score}, is defined as $F=2\times\frac{Precision\times Recall}{Precision+Recall}$. Given the learned action model $\mathcal{M}$ and the reference action model $\mathcal{M}^*$, the bijective function $f_{P\&R}:\mathcal{M} \mapsto \mathcal{M}^*$ is the mapping between the learned and the reference model that maximizes the accumulated {\em F-measure} (considering swaps in the actions with matching headers or parameters with matching types).

%Table~\ref{fig:observationsmap} shows that significantly higher values of {\em precision} and {\em recall} are reported when a learned action schema, $\xi\in\mathcal{M}$, is compared to its corresponding reference schema given by the $f_{P\&R}$ mapping ($f_{P\&R}(\xi)\in \mathcal{M}^*$). The {\em blocksworld} and {\em gripper} domains are perfectly learned from only 25 state observations. These results evidence that in all of the evaluated domains, except for {\em ferry} and {\em satellite}, the learning task swaps the roles of some actions (or parameters) with respect to their role in the reference model.

%The learning scores of several domains in Table~\ref{fig:observationsmap} are above the ones reported in Table~\ref{fig:observationsnomap}. The reason lies in the particular observations comprised by the test sets. As an example, in the {\em driverlog} domain, the action schema {\small \tt disembark-truck} is missing from the learned model because this action is never induced from the observations in the training set; that is, such action never appears in the corresponding \emph{unobserved} plan. The same happens with the {\small \tt paint-down} action of the {\em floortile} domain or {\small \tt move-curb-to-curb} in the {\em parking} domain. Interestingly, these actions do not appear either in the test sets and so the learned action models are not penalized in Table~\ref{fig:observationstest}. Generating {\em informative} and {\em representative} observations for learning planning action models is an open issue. Planning actions include preconditions that are only satisfied by specific sequences of actions, often, with a low probability of being chosen by chance~\cite{fern2004learning}.

>>>>

\subsection{Setup}
We evaluate \FAMA on 14 IPC domains that satisfy the \strips\ requirement~\cite{fox2003pddl2}, all taken from the {\sc planning.domains} repository~\cite{muise2016planning}. Table \ref{tab:domain_features} compiles the features of the tested domains that affect the size of our compilation. From left to right columns report, for each domain, its number of actions, number of predicates, maximum arity of the actions, and maximum arity of the predicates. %and hence, have an impact on the complexity of the learning task. As can be seen in table \ref{tab:domain_features}, the selected domains range from, such as \emph{blocks} and \emph{visitall}, to complex ones, like \emph{floortile} and \emph{satellite}, with higher number of actions and predicates and with higher arity.

@@

\newpage
Figure~\ref{fig:pspace_quality} reports {\em precision} and {\em recall} for the models learned by our approach in the PSPACE-complete scenario (NO action sequence and NO state trajectory). The {\em precision} and {\em recall} values are now lower and they drop as the amount of input knowledge increases. For the PSPACE-complete scenario, {\em precision} and {\em recall} scores have to be taken with a grain of salt because {\bf pure syntax-based metrics are not informative when reformulations appear} (we provide experimental evidence of this in Section \ref{minimal}). Further, the score drop reflects an increasing number of timeouts meaning that no solution is found for several domains within the given time-bound, which hints to the higher worst case complexity of this scenario. Computation time (Figure \ref{fig:pspace_time}) is also higher explained again by this larger number of timeouts.

@@

\subsection{Comparison with \ARMS}
Now we compare the performance of \FAMA to that of \ARMS. \ARMS, as all previous approaches, cannot handle the PSPACE-complete scenario since it requires input plan traces with \FO action sequences so, in this experiment, we will also adopt this assumption.

We define a \emph{degree of observability} $\sigma$ for the state trajectory, ranging from 0\% to 100\%, that measures the probability of observing a literal, and evaluate both approaches as $\sigma$ increases using 5 plan traces as input. When $\sigma = 0$ we have a \NO state trajectory, when $\sigma=100$ we have a \FO state trajectory and all cases in-between correspond to the \PO scenario. Figures \ref{fig:comparison_precision} and \ref{fig:comparison_recall} compare \FAMA and \ARMS in terms of {\em precision} and {\em recall}. Horizontal axes represents the degree of observability, while vertical axes show the average {\em precision} (Figure \ref{fig:comparison_precision}) or {\em recall} (Figure \ref{fig:comparison_recall}). \FAMA dominates in terms of {\em precision} in all cases (the models learned by our approach are 15\% to 24\% more precise than those learned by \ARMS) except for \FO state trajectories. A similar trend is drawn for {\em recall} (Figure \ref{fig:comparison_recall}), where the difference is even larger, meaning that our learned models are more complete. These results evidence the \FAMA capacity to learn better models when few plan traces are available but do not imply that our approach is overall better for NP-complete scenarios (Figure~\ref{fig:np_time} showed that \FAMA cannot handle as much input as \ARMS).

@@

In Table \ref{tab:results_minimum_30_30}, which collects the result of the PSPACE case study with 30\% observability, the scores for some domains are missing. This is the case for {\em floor-tile}, and {\em parking} both considered \emph{puzzle-like} domains, a trait known for putting a strain on the planners. Interestingly enough, {\em hanoi} also qualifies as a \emph{puzzle-like} domain, and we can see how this is reflected on the computation time. With regards to the learned models, we find that they retain similar levels of soundness, while their completeness is lower than in the previous case study. This is specially the case for the preconditions, which have gone from 0.94 recall to 0.54. The explanation for these scores is that input actions are strong constraints that play a key role on how close the learned model are with respect to the GTM. The more these actions are missing the more likely it is for reformulations to occur so pure syntax-based metrics become less relevant.

@@

The effect of missing input actions is more evident in our last case study with \NO action sequences and state trajectories (Table \ref{tab:results_minimum_0_0}). Contrary to what might be expected by looking at the previous table, in this case we are able to find solutions for all domains. The reason for this is that the search is less constrained so there are far more possible solutions for this learning task. This wider range of solutions is also evidenced by the learned models, which in spite of being compliant, are more removed from the GTM. In the table we find that both precision and recall have seen their scores decreased to 0.6 and 0.5, respectively. We argue, however, that syntax-based metrics are not appropriate for scenarios with minimal observability as they cannot cope with the \emph{reformulations} that frequently occur in these circumstances.

To illustrate this, Figure \ref{fig:macroaction} shows the PDDL encoding of the action model of the {\tt stack} operator learned from plan traces with \NO action sequences and state trajectories. This learned action model removes a block from on top of another block and puts it down on the table. There are two main differences with respect to the GTM: (1), the learned action is unstacking a block instead of stacking it and (2), the block ends on the table after the action is applied, working as an {\tt unstack\&putdown} \emph{macro-action}. We refer to the first difference as \emph{role swapping} and it happens when there are missing actions in the plan trace. If no actions are present in the input traces, the names of actions become meaningless (actions are effectively anonymous), meaning that they can interchange their behavior with any other comparable action model. The second issue happens when there are missing states in the input traces, since \emph{macro-actions} apply more than one action in a single step so some intermediate states are skipped.

@@

Table \ref{tab:metric_comparison_100_10} shows the results of the first case study. Looking at the high scores of the syntactic metrics, specially {\em recall}, we can conclude that the learned models are, in fact, fairly similar to the GTM. This supports our conclusion that no reformulation occurs in this case study, which also means that the space of possible solutions is restricted to models close to the GTM. The values for {\em sem-Precision} and {\em sem-Recall} are also high across the table, which is exactly the desired behavior for these metrics given that solutions are close to the GTM. In comparison, both versions of {\em recall} show similar scores, while the difference is bigger for the precision metrics. This means that the semantic version is more lenient towards extra preconditions or effects. This matches what we have observed during the experiments. For instance, the learned model for the {\tt move} action of the {\em hanoi} domain, specifies that both the origin and destination disks must be bigger than the one moving, but the GTM contains only one of these preconditions. This learned model is semantically correct but sintactically different from the GTM and hence, penalized by the {\em precision} metric.

Finally, Table \ref{tab:metric_comparison_0_0} details the results of the \NO/\NO case study. Contrary to the previous case study, the difference between the syntactic and semantic metrics is larger here. Comparing the scores of both versions, we find that learned models that achieved mediocre scores when using the GTM as reference, are in fact reasonable sound and complete, reaching and overall score of 0.91 for both {\em sem-Precision} and {\em sem-Recall}. This means that the models learned by our approach, despite syntactically different from the GTM, require very few editions to explain the testing traces given as input.

@@

Looking at the results of both case studies we can draw two conclusions with regards to the semantic metrics proposed in this paper. The first one is that, when no reformulation occurs, these metrics behave similarly to their syntactic counterparts, which means they are a good substitute when the GTM is not available. The second conclusion is that {\em sem-Precision} and {\em sem-Recall} are better suited to evaluate reformulated models than the original syntactic metrics, since they contemplate valid solutions outside the GTM that successfully explain the given input data.
